Taking Dyads Seriously

### 2018-02-10 paper/1_intro.tex
edited entire manuscript

In this article, we present the Additive and Multiplicative Effects (AME) model, a Bayesian approach for directly modeling relational data to reflect better the interdependencies underlying the data generating process of dyadic data structures \citep{hoff:2008,minhas:etal:2016:arxiv}. We focus on three types of interdependencies that can complicate dyadic analyses. First, dependencies may arise within a set of dyads if a particular actor is more likely to send or receive actions such as conflict.\footnote{In the case of undirected data where there is no clear sender or receiver, it is still essential to take into account the variance in how active actors are in the system.} Additionally, if the event of interest has a clear sender and receiver, we are likely to observe dependencies within a dyad; for example, if a rebel group initiates a conflict with a government, the government will likely reciprocate that behavior. We capture these effects, often referred to as first- and second-order dependencies, respectively, within the additive effects portion of the model. Third order dependencies capture relationships of transitivity, balance, and clusterability between different dyads -- for example, we can only understand why Poland was involved in a dyadic conflict with Iraq in 2003 if we understand that the United States invaded Iraq in 2003 and that Poland often participates in US-led coalitions. The multiplicative effects capture these sorts of dependencies, especially those that result because the specified model has not accounted for some latent set of shared attributes possessed by actors that affect their probability of interacting with one another.

>>>>

In this article, we present the Additive and Multiplicative Effects (AME) model, a Bayesian approach for directly modeling relational data to reflect better the interdependencies underlying the data generating process of dyadic data structures \citep{hoff:2008,minhas:etal:2016:arxiv}. We focus on three types of interdependencies that can complicate dyadic analyses. First, dependencies may arise within a set of dyads if a particular actor is more likely to send or receive actions such as conflict.\footnote{In the case of undirected data where there is no clear sender or receiver, it is still essential to take into account the variance in how active actors are in the system.} Additionally, if the event of interest has a clear sender and receiver, we are likely to observe dependencies within a dyad; for example, if a rebel group initiates a conflict against a government, the government will likely reciprocate that behavior. We capture these effects, often referred to as first- and second-order dependencies, respectively, within the additive effects portion of the model. Third order dependencies capture relationships of transitivity, balance, and clusterability between different dyads--for example, we can only understand why Poland was involved in a dyadic conflict with Iraq in 2003 if we understand that the United States invaded Iraq in 2003 and that Poland often participates in US-led coalitions. The multiplicative effects capture these sorts of dependencies, especially those that result because the specified model has not accounted for some latent set of shared attributes possessed by actors that affect their probability of interacting with one another.

### 2018-02-10 paper/2_model.tex
edited entire manuscript

Scholars typically begin with dyadic data by stacking observations associated with each dyad on top of one another. This makes sense if each observation is independent of the others. For example, a conflict initiated from the United States against Japan is assumed to be independent of any conflictual action that Japan may send to the United States. Additionally, every action sent by Japan to others in the system is considered independent even though each of those interactions involves a common sender, i.e, Japan. While most scholars begin with the assumption that each dyadic interaction is taking place in isolation of the others, we know this assumption to be false both in theory and in practice. Relational data comes with an explicit structure that generally leads to particular types of dependencies. The importance of accounting for the underlying structure of our data has been a lesson well understood, at least, when it comes to time-series cross-sectional data (TSCS) within political science \citep{beck:katz:1995,beck:etal:1998}. As a result, it is now standard practice to take explicit steps to account for the complex data structures that emerge in TSCS applications and the unobserved heterogeneity that they cause.

@@

% hmmm should use an actual IR example here .....
The visualization illustrates that the actors belonging to the same group have a higher likelihood of interacting with each other, whereas interactions across groups are rarer. A prominent example of a network with this type of structure is discussed by \citet{adamic:glance:2005}, who visualizes linkages between right and left leaning political blogs preceding the 2004 United States Presidential Election. \citeauthor{adamic:glance:2005} find that the degree of interaction between right and left leaning blogs was minimal, and that most blogs were linked to others that were politically similar. This showcases the types of higher-order dependencies that can emerge in relational data. First, the fact that interactions were determined by a shared attribute, in this case political ideology, is an example of \textit{homophily}. Homophily explains the emergence of patterns such as transitivity (``a friend of a friend is a friend'') and balance (``an enemy of a friend is an enemy''), which also have a long history in international relations.

>>>>

Scholars working with dyadic data typically begin by stacking observations associated with each dyad on top of one another. This makes sense if each observation is independent of the others. For example, a conflict initiated from the United States against Japan is assumed to be independent of any conflictual action that Japan may send to the United States. Additionally, every action sent by Japan to others in the system is considered independent even though each of those interactions involves a common sender, i.e, Japan. While most scholars begin with the assumption that each dyadic interaction is taking place in isolation of the others, we know this assumption to be false both in theory and in practice. Relational data comes with an explicit structure that generally leads to particular types of dependencies. The importance of accounting for the underlying structure of our data has been a lesson well understood, at least, when it comes to time-series cross-sectional data (TSCS) within political science \citep{beck:katz:1995,beck:etal:1998}. As a result, it is now standard practice to take explicit steps to account for the complex data structures that emerge in TSCS applications and the unobserved heterogeneity that they cause.

@@

% hmmm should use an actual IR example here .....
The visualization illustrates that the actors belonging to the same group have a higher likelihood of interacting with each other, whereas interactions across groups are rarer. A prominent example of a network with this type of structure is discussed by \citet{adamic:glance:2005}, who visualizes linkages between right and left leaning political blogs preceding the 2004 United States Presidential Election. \citeauthor{adamic:glance:2005} find that the degree of interaction between right and left leaning blogs was minimal, and that most blogs were linked to others that were politically similar. This showcases the types of higher-order dependencies that can emerge in relational data. First, the fact that interactions were determined by a shared attribute, in this case political ideology, is an example of \textit{homophily}. Homophily explains the emergence of patterns such as transitivity (``a friend of a friend is a friend'') and balance (``an enemy of a friend is an enemy''), which also have a long history in international relations. For example in International Trade, we might observe homophily wherein states with open economies and an absence of Non-Tariff Barriers are more likely to trade with each-other all else equal; alternatively we might observe heterophily in the sense that states that have greater endowments of capital are more likely to trade with states with greater endowments of land or labor.
